master thesis

### 2018-03-31 Paper - Latex/Paper - Skizze - v4.tex
Thesis - Discussion

\input{arbeit-vorlage-praeambel.tex} % Importiere die Einstellungen aus der Präambel
% hier beginnt der eigentliche Inhalt

@@

\begin{document}
  % Titelseite
  \begin{titlepage}
    \pagestyle{empty}
  	\begin{center}
      {\Large Freie Universität Berlin} \\
    	\begin{Huge}
      	Fachbereich Mathematik und Informatik \\
      	\vspace{3mm}
    	\end{Huge}
    	\vspace{2cm}
    	\begin{Large}
        \textbf{An analysis of maximum parsimony algorithms to predict parasitism in Eukaryota} \\
        \vspace{3mm}
        using a large multifurcated phylogenetic synthesis tree \\
    	\end{Large}
      \vspace{3cm}
      \textbf{Submitted on:} \\
      3 April 2018 \\
    	\vspace{2cm}
    	Lydia Buntrock \\
      E-Mail: info@irallia.de \\
     	\vspace{3cm}
      \textbf{Supervisor:} \\
      Prof. Dr. Bernhard Y. Renard \\
      \& \\
      Prof. Dr. rer. nat. Emanuel Heitlinger \\      
  	\end{center}
    \clearpage
    \pagenumbering{Roman}
  \end{titlepage}

@@

  Parasitism can be defined as an interaction between species in which one of the interaction 
    partners, the parasite, lives in or on the other, the host. The parasite draws food from its 
    host and harms it in the process. According to estimates, above 40\% of all eukaryotes are 
    parasites. Nevertheless, it is computationally difficult to obtain information whether a 
    particular taxon is a parasite making it difficult to query large sets of taxa.

@@

  This represents, to our knowledge, the first comprehensive computational reconstruction of the 
    emergence of parasitism in eukaryotes. We argue that such an approach is necessary to allow 
    further incorporation of parasitism as an important trait in species interaction databases and 
    in individual studies on eukaryotes e.g. in the microbiome.

@@

  For these large phylogenetic synthesis trees, however, ancestral state reconstruction has so far 
    only been done for Bacteria and Archaea for binary traits by Goberna and Verdú \cite{Goberna2015}.
    However, this differs from eukaryotes in the sense that complex traits such as parasitism depend
    on more than one gene.

@@

  The objectives of this work are the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Accomplish reconstructing on the Eucaryotic synthesis tree of OTL. \\
  The goal of Point 1 is to evaluate the possible methods based on a simulation of our data 
    situation. \\
  The Sankoff algorithm implemented by Louca et al. is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states 
    with the aid of this algorithm. And then perform an evaluation of the results. \\

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Aims}
  The aim of this thesis is the application of maximum parsimony algorithms to non-binary trees and 
    very large data sets to predict unknown leaf node states. \\
  We discuss different ancestral state reconstruction methods and test the most appropriate ones on 
    our data. Hereby, we want to find out how far it is possible to use the Open Tree of Life (OTL) 
    to expandavailable information via phylogenetic trait prediction. As a basis for known states, 
    we use the Global Biotic Interactions (GloBI) database, which cathegorises 34,860 and 25,962 
    species as free-living and parasites. This data is used to perform an ancestral state 
    reconstruction and predict states for over $\sim$2.3 million (97.34\%) leaf nodes without state 
    information.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  In this thesis, a maximum parsimony algorithm is applied to the Eukaryota tree to obtain an 
    ancestral state reconstruction of free-living versus parasite states. \\
  This chapter is divided into the following sections: the description of the data sources 
    (section \ref{sec:methods - data sources}) used and the analysis of these data as a preparation 
    for the simulation (section \ref{sec:methods - metadata analysis}). A description of the 
    analyzed methods for the ancestral state reconstruction (section 
    \ref{sec:methods - ancestral state reconstruction}) and then an explanation of how they fit the 
    problem at hand (section \ref{sec:methods - simulation}). And at the end, the real data analysis 
    with the Sankoff method (section \ref{sec:methods - real data analysis}). \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulting algorithm on the original data.
      (7) Evalute and interprete the results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

@@

    Furthermore the Open Tree Taxonomy (OTT) from OTL is used because it includes most of the known 
      taxonomies and is synthesised by prefering taxonomies that match with available phylogenetic 
      data. The team from OTL prefer a maximum number of species \cite{Hinchliff2015}, this results 
      in the synthesis between taxonomy and phylogeny.

    For the state information the Global Biotic Interactions database (GloBI) is taken 
      \cite{Poelen2014} (downloaded on 29.01.2018). \\
    This database consists of entries of the form: species A (source) interacts with B (target). \\
    A number of interactions have been identified\footnote{\hyperlink{
      https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
      }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}; Last checked: 22.03.2018.}
      , including those indicating whether the species source or target has become a parasite or a 
      free-living species from the biological perspective. They are the following:
    \begin{itemize}
      \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
        hasParasite, hostOf
      \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
      \item parasite source: parasiteOf, pathogenOf
      \item parasite target: hasParasite, hasPathogen
    \end{itemize}
    Of these interactions, e.g. species A parasitize species B, the state of the species is 
      determined, here is species A parasitic and species B free-living. In the case that a parasite 
      conquers (parasitizes) another parasite yields conflicting states for the second species. 
      This is solved by preferring parasitic. \\
    For each species known identifyers are stored in GloBI. This includes OTT (the taxonomy of OTL). 
      All species that have stored an OTT indentifyer and have a matching interaction are divided into 
      two lists: parasites and free-livings.

@@

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (e.g. \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}).

@@

      (The original Fitch algorithm is designed to minimize transitions without predicting actual states 
        of internal nodes, so it is just the first traversal.) \\
        \begin{wrapfigure}[20]{r}{0.5\textwidth}
          \centering
          \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees. \\
            The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
            root node) consists of three sets, where the last set is the final one (bold). \\
            From the second internal node (seen from the root node) there are several possibilities to 
            create the second and third set.}
          \label{fig: binary Fitch}
        \end{wrapfigure}
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the incision or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that are all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The parent node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item If there are several siblings, is the cut or union made first of these, or is everything 
          made directly including the parent node.
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the parent node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch are formed:
      \begin{enumerate}
        \item Fitch 1: First state set of parent node; intersection/union of siblings first.
        \item Fitch 2: First state set of parent node; intersection/union of siblings together with parent node.
        \item Fitch 3: Both state sets of parent node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of parent node; intersection/union of siblings together with parent node sets.
      \end{enumerate}

@@

    The next step is to simulate states for all nodes. \\
    Here the influence of the biological parameters as transition probabilities and distributions of
      \begin{wrapfigure}[21]{r}{0.55\textwidth}
        \centering
        \includegraphics[width=0.55\textwidth]{Figures/40-60_all.jpg}
        \caption{Set node states: Distribution of states (left); traversion through the tree (right). \\
          Start with a free-living root node (FL: red). \\
          (1) + (2) Draw random numbers for its children from the free-living distribution (red), the numbers are under the threshold 
          $\rightarrow$ the nodes are again free-living; go on with the children. \\
          (3) The number drawn is above the threshold. $\rightarrow$ The node state changes to parasitic (P: blue). \\
          (4) Now draw random numbers from the parasite distribution (red) until one number lies unter the threshold. Then change back.}
        \label{fig:set node states}
      \end{wrapfigure} 
      states is included. \\
    Since there is no statement about general transition probabilities, these are all set the same: \\
    $\mathcal{P}(FL \rightarrow P) = \mathcal{P}(P \rightarrow FL)$.

@@

    After traversing through the tree, each state is saved in a nodelist associated with the node 
      indentifyer which is the OTT from OTL.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Results}
  This work deals with the ancestral state reconstruction of the entire Eukaryota relatives tree.

@@

    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l||r|r| }
          \hline
          \bfseries Subtree of & \bfseries Unknown States & \bfseries Multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34\%  & 89.45\% \\
          \hline \hline
          Metazoa         & 96.44\%  & 87.93\% \\ \hline
          Fungi           & 98.87\%  & 96.97\% \\ \hline
          Chloroplastida  & 99.14\%  & 89.46\% \\
          \hline \hline            
          % Arthropoda      & 97.49\%  & 89.95\% \\ \hline
          Apicomplexa     & 86.26\%  & 87.16\% \\ \hline
          Nematoda        & 89.01\%  & 88.59\% \\ \hline
          Chordata        & 88.59\%  & \cellcolor{green!50}66.49\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73\%  & 80.34\% \\
          \hline \hline            
          Insecta         & 97.11\%  & 90.78\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\
        The percentage values show the proportion of missing information of: unknown states (missing state 
          information of leaf nodes) and multifurcation (missing internal nodes). \\
        The subtrees are from different taxa: domain (Eukaryota), kingdom (Metazoa, Fungi, 
          Chloroplastida), phylum (Apicomplexa, Nematoda, Chordata, Platyhelminthes) and class 
          (Insecta). \\
        The two by far smallest values are highlighted in green.} 
      \label{table:percentage loss information subtrees} 
    \end{table}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Comparison of different models for missing state information}
      Next to the problem of the multifurcation of the tree is the little interaction data thet we have 
        for the species. For the ancestral state reconstruction, we need information about the states 
        (free-living or parasite) in the leaf nodes. \\
      The eukaryotic synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we got 51,337 free-living species and 47,332 parasite species 
        for the whole tree of life. From the Eukaryota we could determine 25,962 and 34,860 species as 
        parasites and free-living. With 2,293,463 leaf nodes we still have about 97.34\% unknown leaf 
        nodes.

@@

    Since we have a lot of missing data in most subtrees (both internal nodes and state informations), 
      the Sankoff gives a better prediction and is thus used for real data analysis.

@@

      We examined the influence of the omitted data on the prediction. On average, about twice as many 
        leaf nodes are predicted differently. %\color{red}The variance is very high. This means that 
        % we have a high degree of dispersion and thus a stochastic situation exists. It also describes 
        % the width of the present probability function. The standard deviation is about five times as 
        % high as the number of omitted nodes, so the variability is quintupled. \color{black} 
        Table \ref{table:statistics cross-validation} shows these results. \\
      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |cl||r|r|r|r|r| }
            \hline
            & & \bfseries min & \bfseries max & \bfseries mean & \bfseries variance ($\sigma^2$) & $\sigma$ \\
            \hline \hline
            \multirow{3}{*}{\bfseries distance} & \bfseries all     & 0 & 3587.70 & 224.96 & 313650.61 & 560.05 \\
            & \bfseries leaf nodes                                  & 0 & 3021.12 & 208.69 & 248103.38 & 498.10 \\
            & \bfseries internal nodes                              & 0 & 566.58 & 16.28 & 4927.95 & 70.20 \\ \hline
            \multirow{3}{*}{\bfseries lost} & \bfseries all states  & 100 & 100 & 100 & 0 & 0 \\
            & \bfseries FL states                                   & 44 & 66 & 57.25 & 19.50 & 4.42 \\
            & \bfseries P states                                    & 34 & 56 & 42.75 & 19.50 & 4.42 \\
            \hline
          \end{tabular}
        \end{center}
        \caption{Statistics about the leave-100-out cross-validation \\
          The distance between original and new states is calculated using the Euclidean metric. This 
            is summed over all states / all leaf node states and all internal node states. \\
          The lower half of the table describes the distribution of the 'lost' states between parasites 
            (P) and free-livings (FL).}
        \label{table:statistics cross-validation}
      \end{table}

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Discussion}

To analyze the results, we have selected some phyla (subtrees) to evaluate our results 
selectively from the biological point of view: Chordata, Nematoda, Platyhelminthes and 
Apicomplexa. \\

Several factors such as the distribution of existing input data via parasitics and free-livings, 
faulty input data from GloBI and reinforcement of errors by multifurcation play an important role which becomes evident in these examples. \\

An important factor here is the credibility of the results. The accuracy of the input data 
stands and falls with the presence and the correctness of the data of GloBI. Errors of incorrect 
input data can be amplified by incorrect prediction of unknown species and can be reversed in 
order to improve the data situation of GloBI. \\
Since we look at such large trees we can not expect to know all the parasites, so we look at 
individual positives. They are positive in the sense that the majority have the opposite state.

In contrast to the other phyla examined, the phylogeny in the Chordata is more pronounced (less 
        multifurcation) (see table \ref{table:percentage loss information subtrees}). This results in 
        less variance of errors. What is reflected in the results shown in table 
        \ref{table:results of some selected taxa}. There are 18 parasites as input data and only 8 more are 
        predicted. The Chordata mostly consist of free-living species, so this seems believable. We 
        started with 99.83\%  species and predict 99.97\% species as free-living (including already 
        known nodes).

      We mapped the few parasitic species in a rough taxonomy (see Figure \ref{fig:ChordataParasites}): \\
      Known parasitic birds belong to the order Sauria. Here we know from Rothschild here that are 
        breeding parasites, like the cuckoo and clepto-parasites as the skuas \cite{Rothschild1957}. We
        got 6 input parasites from GloBI and there are no predictions: A woodpecker - 
        \textit{Sphyrapicus varius} and a duck - \textit{Aix sponsa}, a cow bird - \textit{Molothrus ate} 
        known as broodparasite and some others. \\
      An example of the amplification of mistakes here are the carp. There is a paper from which GloBI 
        concludes: Grass carp (\textit{Ctenopharyngodon idella)} has Pathogen common carp 
        (\textit{Cyprinus carpio})\footnote{
          \hyperlink{https://www.globalbioticinteractions.org/?interactionType=hasParasite&targetTaxon=Cyprinus\%20carpio}
          {https://www.globalbioticinteractions.org/?interactionType=hasParasite\&targetTaxon=Cyprinus\%20carpio}
        }. Since there is hardly any information about free-living, it follows that all siblings are 
        also predicted to be parasitic. \\
        % Other birds: Columba livia, Nestor meridionalis, Sphyrapicus varius \\
        % Snake: Storeria dekayi (probably caused by wrong input data.)
      % \begin{itemize}
      %   \item Order Acanthomorphata(3):, one slime fish (family Blenniidae) and two others without family
      %   \item Order Anura (1): a toad (Family Bufonidae) - there exists some parasitic species in here, 
      %     they eat fin or pieces of larger fish.
      %   \item Order Siluriformes (1): a catfish (Family Loricariidae) - there exists some parasitic 
      %     species in here.
      % \end{itemize}

      The Apicomplexa are a parasitic phylum. We found only one input organism: Stemonitis fusca as a 
        free-living species. In GloBI it is listed as being parasitized by Nectria candicans and Nectriopsis 
        sporangiicola\footnote{
          \hyperlink{https://www.globalbioticinteractions.org/?interactionType=parasiteOf&targetTaxon=Stemonitis\%20fusca}
          {https://www.globalbioticinteractions.org/?interactionType=parasiteOf\&targetTaxon=Stemonitis\%20fusca}
        }. The algorithm has not predicted new free-livings. \\

>>>>

\input{arbeit-vorlage-praeambel.tex}

@@

\begin{document}
  \begin{titlepage}
    \pagestyle{empty}
  	\begin{center}
      {\Large Freie Universität Berlin} \\
    	\begin{Huge}
      	Fachbereich Mathematik und Informatik \\
      	\vspace{3mm}
    	\end{Huge}
    	\vspace{2cm}
    	\begin{Large}
        \textbf{An analysis of maximum parsimony algorithms to predict parasitism in Eukaryota} \\
        \vspace{3mm}
        using a large multifurcated phylogenetic synthesis tree \\
    	\end{Large}
      \vspace{3cm}
      \textbf{Submitted on:} \\
      3 April 2018 \\
    	\vspace{2cm}
    	Lydia Buntrock \\
      E-Mail: info@irallia.de \\
     	\vspace{3cm}
      \textbf{Supervisor:} \\
      Prof. Dr. Bernhard Y. Renard \\
      \& \\
      Prof. Dr. rer. nat. Emanuel Heitlinger \\      
  	\end{center}
    \clearpage
    \pagenumbering{Roman}
  \end{titlepage}

@@

  Parasitism can be defined as an interaction between species in which one of the interaction 
    partners, the parasite, lives in or on the other, the host. The parasite draws food from its 
    host and harms it in the process. According to estimates, above 40\% of all Eukaryota are 
    parasites. Nevertheless, it is computationally difficult to obtain information whether a 
    particular taxon is a parasite making it difficult to query large sets of taxa.

@@

  This represents, to our knowledge, the first comprehensive computational reconstruction of the 
    emergence of parasitism in Eukaryota. We argue that such an approach is necessary to allow 
    further incorporation of parasitism as an important trait in species interaction databases and 
    in individual studies on Eukaryota e.g. in the microbiome.

@@

  For these large phylogenetic synthesis trees, however, ancestral state reconstruction has so far 
    only been done for Bacteria and Archaea for binary traits by Goberna and Verdú \cite{Goberna2015}.
    However, this differs from Eukaryota in the sense that complex traits such as parasitism depend
    on more than one gene.

@@

  The objectives of this work are the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Accomplish reconstructing on the Eukaryota synthesis tree of OTL. \\
  The goal of Point 1 is to evaluate the possible methods based on a simulation of our data 
    situation. \\
  The Sankoff algorithm implemented by Louca et al. is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states 
    with the aid of this algorithm. And then perform an evaluation of the results. \\

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Aims}
  The aim of this thesis is the application of maximum parsimony algorithms to non-binary trees and 
    very large data sets to predict unknown leaf node states. \\
  We discuss different ancestral state reconstruction methods and test the most appropriate ones on 
    our data. Hereby, we want to find out how far it is possible to use the Open Tree of Life (OTL) 
    to expand available information via phylogenetic trait prediction. As a basis for known states, 
    we use the Global Biotic Interactions (GloBI) database, which categorises 34,860 and 25,962 
    species as free-living and parasites. This data is used to perform an ancestral state 
    reconstruction and predict states for over $\sim$2.3 million (97.34\%) leaf nodes without state 
    information.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  In this thesis, a maximum parsimony algorithm is applied to the Eukaryota tree to obtain an 
    ancestral state reconstruction of free-living versus parasite states. \\
  This chapter is divided into the following sections: the description of the data sources 
    (section \ref{sec:methods - data sources}) used and the analysis of these data as a preparation 
    for the simulation (section \ref{sec:methods - metadata analysis}). A description of the 
    analyzed methods for the ancestral state reconstruction (section 
    \ref{sec:methods - ancestral state reconstruction}) and then an explanation of how they fit the 
    problem at hand (section \ref{sec:methods - simulation}). And at the end, the real data analysis 
    with the Sankoff method (section \ref{sec:methods - real data analysis}). \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realistic simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulting algorithm on the original data.
      (7) Evaluate and interpret the results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

@@

    Furthermore the Open Tree Taxonomy (OTT) from OTL is used because it includes most of the known 
      taxonomies and is synthesised by preferring taxonomies that match with available phylogenetic 
      data. The team from OTL prefer a maximum number of species \cite{Hinchliff2015}, this results 
      in the synthesis between taxonomy and phylogeny.

    For the state information the Global Biotic Interactions database (GloBI) is taken 
      \cite{Poelen2014} (downloaded on 29.01.2018). \\
    This database consists of entries of the form: species A (source) interacts with B (target). \\
    A number of interactions have been identified\footnote{\hyperlink{
      https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
      }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}; Last checked: 22.03.2018.}
      , including those indicating whether the species source or target has become a parasite or a 
      free-living species from the biological perspective. They are the following:
    \begin{itemize}
      \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
        hasParasite, hostOf
      \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
      \item parasite source: parasiteOf, pathogenOf
      \item parasite target: hasParasite, hasPathogen
    \end{itemize}
    Of these interactions, e.g. species A parasitize species B, the state of the species is 
      determined, here is species A parasitic and species B free-living. In the case that a parasite 
      conquers (parasitizes) another parasite yields conflicting states for the second species. 
      This is solved by preferring parasitic. \\
    For each species known identifiers are stored in GloBI. This includes OTT (the taxonomy of OTL). 
      All species that have stored an OTT identifier and have a matching interaction are divided into 
      two lists: parasites and free-livings.

@@

      Input: A rooted, binary tree, with state information in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (e.g. \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}).

@@

      (The original Fitch algorithm is designed to minimize transitions without predicting actual states 
        of internal nodes, so it is just the first traversal.) \\
        \begin{wrapfigure}[20]{r}{0.5\textwidth}
          \centering
          \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees. \\
            The unknown leaf node is described with both states. Computed internal nodes (exclusive the 
            root node) consists of three sets, where the last set is the final one (bold). \\
            From the second internal node (seen from the root node) there are several possibilities to 
            create the second and third set.}
          \label{fig: binary Fitch}
        \end{wrapfigure}
      The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
        more than two children may be present for the first traversal, but the intersection or union may 
        also be formed over more than two sets. Also in the second traversing, there may be several 
        sibling nodes. However, there are several possibilities here that are all tested and compared 
        in the simulation. Some of these options are already available in the binary case:
      \begin{itemize}
        \item The parent node has (except for the root node) two state sets, because he came through 
          the up-traversing previously. Are both sets used or only the first traversing?
        \item If there are several siblings, is the cut or union made first of these, or is everything 
          made directly including the parent node.
      \end{itemize}
      The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
        both possibilities of the three sets. \\
      Cunningham uses only the first state set of the parent node \cite{Cunningham1998}. \\
      From these two points four different versions of Fitch are formed:
      \begin{enumerate}
        \item Fitch 1: First state set of parent node; intersection/union of siblings first.
        \item Fitch 2: First state set of parent node; intersection/union of siblings together with parent node.
        \item Fitch 3: Both state sets of parent node; intersection/union of siblings first.
        \item Fitch 4: Both state sets of parent node; intersection/union of siblings together with parent node sets.
      \end{enumerate}

@@

    The next step is to simulate states for all nodes. \\
    Here the influence of the biological parameters as transition probabilities and distributions of
      \begin{wrapfigure}[21]{r}{0.55\textwidth}
        \centering
        \includegraphics[width=0.55\textwidth]{Figures/40-60_all.jpg}
        \caption{Set node states: Distribution of states (left); traversion through the tree (right). \\
          Start with a free-living root node (FL: red). \\
          (1) + (2) Draw random numbers for its children from the free-living distribution (red), the numbers are under the threshold 
          $\rightarrow$ the nodes are again free-living; go on with the children. \\
          (3) The number drawn is above the threshold. $\rightarrow$ The node state changes to parasitic (P: blue). \\
          (4) Now draw random numbers from the parasite distribution (red) until one number lies under the threshold. Then change back.}
        \label{fig:set node states}
      \end{wrapfigure} 
      states is included. \\
    Since there is no statement about general transition probabilities, these are all set the same: \\
    $\mathcal{P}(FL \rightarrow P) = \mathcal{P}(P \rightarrow FL)$.

@@

    After traversing through the tree, each state is saved in a nodelist associated with the node 
      identifier which is the OTT from OTL.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Results}
  In the first part of this work, we were able to provide the proof of concept for the Sankoff 
    algorithm to perform an ancestral state reconstruction of the present Eukaryota tree. We will 
    therefore discuss in the following how significant the result of this reconstruction is.

    To do so, we first look at the input data we have: the use of Open Tree of Life for the tree 
      gives us the greatest approximation to a phylogeny of the Eukaryota. Since such a large 
      phylogeny has not been completely dissected, this is only a synthesis of phylogenetic trees 
      with a taxonomy as a backbone. As a result, the tree is far from binary, we found a 
      multifurcation of 89.45 \%. However, this is highly variable and therefore the reconstruction 
      is still very good on wide parts of the tree. \\
    On the other hand, we used the Global Biotic Interactions (GloBI) database to get the state 
    information for the leaf nodes. However, we were able to extract only 2.66\% of the node 
    information from GloBI. However, we found out that there are more data in GloBI that could be 
    taken away with some work.

    % One last limitation also exists in the data situation itself. We have pointed out in various 
    %   sections that we have a strong multifurcation and only a few data about leaf node states.

    Next, let's consider the limitations of the simulation that led to the choice of the Sankoff 
      algorithm. Our own main point of criticism is that we did not simulate the transition 
      probabilities and estimate an unrealistically high number of transitions (origins and losses). 
      The Sankoff algorithm is very well scored despite this increased difficulty (more transitions 
      means more complex predictions) and its proven feasibility.

    Finally, we analyze the validation of our ancestral state reconstruction results. We will do a 
      selective analysis from a biological perspective and discuss the leave-100-out 
      cross-validation. \\
    Above all, the 'biological view' gives a first impression of the credibility of the results. But 
      it can not make any statistical statements. Here the interested researcher may have to analyze 
      the subtrees of being relevant to him for himself. \\
    Of course, cross-validation can be expanded. With more computing power, a leave-1-out 
      cross-validation could be performed. However, our result on this point is 98.17\% correct. \\

\todo{old:}

  This work deals with the ancestral state reconstruction of the entire Eukaryota relatives tree.

@@

    \begin{table}[h!]
      \begin{center}
        \begin{tabular}{ |l||r|r| }
          \hline
          \bfseries Subtree of & \bfseries Unknown States & \bfseries Multifurcation \\ 
          \hline \hline
          Eukaryota       & 97.34\%  & 89.45\% \\
          \hline \hline
          Chloroplastida  & 99.14\%  & 89.46\% \\ \hline
          Fungi           & 98.87\%  & 96.97\% \\ \hline
          Metazoa         & 96.44\%  & 87.93\% \\
          \hline \hline
          Apicomplexa     & 86.26\%  & 87.16\% \\ \hline
          Arthropoda      & 97.49\%  & 89.95\% \\ \hline
          Chordata        & 88.59\%  & \cellcolor{green!50}66.49\% \\ \hline
          Nematoda        & 89.01\%  & 88.59\% \\ \hline
          Platyhelminthes & \cellcolor{green!50}68.73\%  & 80.34\% \\
          \hline \hline            
          Insecta         & 97.11\%  & 90.78\% \\
          \hline  
        \end{tabular}
      \end{center}
      \caption{Examination of subtrees regarding missing information. \\
        The percentage values show the proportion of missing information of: unknown states (missing state 
          information of leaf nodes) and multifurcation (missing internal nodes). \\
        The subtrees are from different taxa: domain (Eukaryota), kingdom (Metazoa, Fungi, 
          Chloroplastida), phylum (Apicomplexa, Nematoda, Chordata, Platyhelminthes) and class 
          (Insecta). \\
        The two by far smallest values are highlighted in green.} 
      \label{table:percentage loss information subtrees} 
    \end{table}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{Comparison of different models for missing state information}
      Next to the problem of the multifurcation of the tree is the little interaction data thet we have 
        for the species. For the ancestral state reconstruction, we need information about the states 
        (free-living or parasite) in the leaf nodes. \\
      The Eukaryota synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we got 51,337 free-living species and 47,332 parasite species 
        for the whole tree of life. From the Eukaryota we could determine 25,962 and 34,860 species as 
        parasites and free-living. With 2,293,463 leaf nodes we still have about 97.34\% unknown leaf 
        nodes.

@@

    Since we have a lot of missing data in most subtrees (both internal nodes and state information), 
      the Sankoff gives a better prediction and is thus used for real data analysis.

@@

      We examined the influence of the omitted data on the prediction. On average, about twice as many 
        leaf nodes are predicted differently. %\color{red}The variance is very high. This means that 
        % we have a high degree of dispersion and thus a stochastic situation exists. It also describes 
        % the width of the present probability function. The standard deviation is about five times as 
        % high as the number of omitted nodes, so the variability is quintupled. \color{black} 
        Table \ref{table:statistics cross-validation} shows these results. \\
      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |cl||r|r|r|r|r| }
            \hline
            & & \bfseries min & \bfseries max & \bfseries mean & \bfseries variance ($\sigma^2$) & $\sigma$ \\
            \hline \hline
            \multirow{3}{*}{\bfseries distance} & \bfseries all     & 0 & 3587.70 & 224.96 & 313650.61 & 560.05 \\
            & \bfseries leaf nodes                                  & 0 & 3021.12 & 208.69 & 248103.38 & 498.10 \\
            & \bfseries internal nodes                              & 0 & 566.58 & 16.28 & 4927.95 & 70.20 \\ \hline
            % \multirow{3}{*}{\bfseries lost} & \bfseries all states  & 100 & 100 & 100 & 0 & 0 \\
            % & \bfseries FL states                                   & 44 & 66 & 57.25 & 19.50 & 4.42 \\
            % & \bfseries P states                                    & 34 & 56 & 42.75 & 19.50 & 4.42 \\
            % \hline
          \end{tabular}
        \end{center}
        \caption{Statistics about the leave-100-out cross-validation \\
          The distance between original and new states is calculated using the Euclidean metric. This 
            is summed over all states, all leaf node states and all internal node states.}
          % The lower half of the table describes the distribution of the 'lost' states between parasites 
            % (P) and free-livings (FL).}
        \label{table:statistics cross-validation}
      \end{table}

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Discussion}
  This chapter deals with the question of how trustworthy our result of the ancestral state 
    reconstruction of the Eukaryota tree is. And further on how well our simulation can simulate the 
    real problem and thus make statements about the predictive power of the applied Sankoff algorithm.
